---
title: "Nanobodies Sequencing Analysis"
author:
  - name: "Francesco Lescai"
    affiliation: "Università di Pavia"
    email: "francesco.lescai@unipv.it"
  - name: "Davide Bagordo"
    affiliation: "Università di Pavia"
    email: "davide.bagordo@unipv.it"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    theme: default
    toc: true
editor:
  options:
    chunk-output-type: console
params:
  clusterList: null
  histoList: null
  tableList: null
  sampleData: null
  sizeThreshold: null
  loopFile: null
  calcTree: null
---


```{r}
#| label: library installation
#| include: false

# Carica le librerie necessarie per l'analisi e imposta le opzioni di knitr per il rendering dei chunk di codice.
library(knitr)
library(ggplot2)
library(tidyverse)
library(kableExtra)
library(ggtree)
library(Biostrings)
# library(msa)
library(plotly)
library(seqinr)
library(ape)

opts_chunk$set(
    message = FALSE,
    error = FALSE,
    warning = FALSE,
    tidy = TRUE,
    tidy.opts = list(blank = FALSE),
    cache = TRUE,
    cache.lazy = FALSE,
    echo = FALSE,
    results = "asis",
    fig.pos = "H",
    fig.wide = TRUE
)
```

```{r }
#| label: Data Preparation

# Questa sezione legge i parametri passati al documento, tra cui le liste di file di cluster, istogrammi e tabelle, oltre ai dati campione e altri parametri di configurazione.

clusterfiles <- unlist(strsplit(params$clusterList, ","))
histofiles <- unlist(strsplit(params$histoList, ","))
tablefiles <- unlist(strsplit(params$tableList, ","))
sampleMetaData <- read_tsv(params$sampleData)
sizeThreshold <- unlist(params$sizeThreshold)
loopFile <- unlist(params$loopFile)
dotree <- ifelse(unlist(params$calcTree) == "true", TRUE, FALSE)
```

# Introduction

In this analysis, we aim to generate the total number of different nanobody CDR3s that exist in a library, and the relative abundance of each. We then perform further analysis to show which B cell lineages were affinity matured. For our initial analysis we follow some of the steps used by Deschaght et al (2017).

# Results

In the first section, we briefly summarise the characteristics of the clusters. In the second one, we look more in detail at the CDR3 in the libraries.

## Clusters

In the following, we review the distribution of members of clusters and their sequence identity to the representative.

```{r}
#| label: ClusterMembership

# Legge i dati di cluster da ciascun file nella lista e li combina in un unico dataframe.

clusterbig <- data.frame()
for (file in clusterfiles) {
    samplename <- gsub("_clusters.summary", "", file)
    writeLines(paste0("-------> received sample name as = ", samplename))
    clusterdata <- read_csv(file)
    head(clusterdata)
    clusterdata$Sample <- samplename
    clusterbig <- rbind(clusterbig, clusterdata)
    rm(clusterdata)
}
```

```{r}
#| label: ClusterSizePlot

# Genera un istogramma delle dimensioni dei cluster.

ggplot(data = clusterbig) +
    geom_histogram(aes(x = Count)) +
    scale_x_continuous(limits = c(0, 500)) +
    scale_y_continuous(limits = c(0, 5000)) +
    xlab("Number of Cluster Members") +
    ggtitle("Cluster Sizes") +
    facet_wrap(~Sample)
```

```{r}
#| label: clusterMembersPlot

# Genera un grafico a dispersione che mostra l'identità della sequenza rispetto al rappresentante del cluster.
# I cluster singleton hanno Identity = NA (nessun membro da confrontare) e vengono esclusi.

ggplot(data = clusterbig %>% filter(!is.na(Identity))) +
    geom_point(aes(x = Count, y = Identity), position = "jitter") +
    xlab("Number of Cluster Members") +
    ylab("Mean Sequence Identity of Members\nto Cluster Representative (%)") +
    facet_wrap(~Sample)
```

We summarise cluster counts based on the number of member sequences.

```{r}
#| label: clustercountsTable

# Crea una tabella che riassume il numero di cluster in base alla dimensione del cluster.
# Total_sequences e' il denominatore necessario per calcolare le frazioni clonali
# (quota della libreria occupata dal clone dominante e dai primi dieci cloni).

clusterbig$five <- clusterbig$Count >= 5
clusterbig$hundred <- clusterbig$Count >= 100
clusterbig$thousand <- clusterbig$Count >= 1000

clustercounts <- clusterbig %>%
    group_by(Sample) %>%
    summarize(
        Clusters = n(),
        Clusters_of_5 = sum(five),
        Clusters_of_100 = sum(hundred),
        Clusters_of_1000 = sum(thousand),
        Total_sequences = sum(Count),
        Top1_count = max(Count),
        Top10_count = sum(sort(Count, decreasing = TRUE)[1:min(10, n())])
    )

kable(clustercounts) %>%
    kable_styling()
```

## CDR3 analysis

The CDR3 of each cluster representative is annotated with nanoCDR-X (Bagordo et al. 2026). The reference definition is CDR3-IMGT, positions 105–117 of the IMGT unique numbering, between the conserved cysteine 104 and tryptophan 118, both excluded (Lefranc et al. 2003); the agreement of the nanoCDR-X calls with it is reported in the CDR3 annotation QC of the interactive report. We also compute the length distribution of these CDR3 sequences.

```{r }
#| label: cdrhistogram

# Legge i dati degli istogrammi CDR3 da ciascun file e li combina in un unico dataframe.

cdrhists <- data.frame()
for (file in histofiles) {
    samplename <- gsub("_cdr3.hist", "", file)
    histdata <- read.csv(file, header = TRUE)
    colnames(histdata)[1:2] <- c("Size", "Count")
    histdata$Sample <- rep(samplename, times = nrow(histdata))
    cdrhists <- rbind(cdrhists, histdata)
}
```

```{r}
#| label: CDR3countsTable

# Crea una tabella che riassume il numero di CDR3 unici per campione.

cdrcounts <- cdrhists %>%
    group_by(Sample) %>%
    summarize(Unique_CDR3s = sum(Count))

kable(cdrcounts) %>%
    kable_styling()
```

```{r}
#| label: CDR3sizeplot

# Genera un istogramma della distribuzione delle dimensioni delle sequenze CDR3 uniche.

ggplot(data = cdrhists) +
    geom_histogram(aes(x = Size, y = Count), stat = "identity") +
    facet_wrap(~Sample) +
    ggtitle("Size Distribution of Unique CDR3s")
```

## Evolution of CDR3

To be completed. Currently, only outputting a table file.

```{r }
#| label: fastaRep

# Legge i dati delle sequenze fasta da ciascun file e li combina in un unico dataframe.

fastaSeq <- data.frame()
for (fasta in tablefiles) {
    sample <- gsub("_cdr3.tsv", "", fasta)
    seqs <- read_tsv(fasta)
    fastaSeq <- rbind(
        fastaSeq,
        cbind(seqs,
            sample = sample
        )
    )
    rm(seqs)
}
```

```{r}
#| label:  mergeData

# Unisce i dati dei cluster con i dati delle sequenze fasta.

boostEvol <- clusterbig
    clusterbig$Representative <- as.character(clusterbig$Representative)
    fastaSeq$ID <- as.character(fastaSeq$ID)

boostEvol <- clusterbig %>%
    left_join(fastaSeq,
        by = c(
            "Representative" = "ID",
            "Sample" = "sample"
        )
    )
```

temporarily taken from merged dataset, otherwise would be loaded from file.

```{r}
#| label: sampleMetaData

# Unisce i dati risultanti con i metadati del campione e salva il risultato in un file TSV.

boostEvol <- boostEvol %>%
    left_join(sampleMetaData,
        by = c("Sample" = "ID")
    )
write_tsv(boostEvol, "cdr3_boost_overview_table.tsv")
```

```{r}
#| label: MSAandTreeStore

# Questa sezione esegue l'allineamento multiplo delle sequenze CDR3 e genera un albero filogenetico per ciascun individuo.

individualData <- list()
treeDataList <- list()
treeMetaDataList <- list()
treePlotList <- list()

for (ind in unique(boostEvol$individual)) {
    tmpData <- boostEvol %>%
        filter(individual == ind) %>%
        filter(!is.na(CDR3)) %>%
        filter(Count > sizeThreshold) %>%
        group_by(CDR3) %>%
        summarise(
            clusterSize = max(Count),
            representativeSeq = Representative[which(Count == max(Count))],
            individual = individual[which(Count == max(Count))],
            sample = Sample[which(Count == max(Count))],
            immunisation = immunisation[which(Count == max(Count))],
            boost = boost[which(Count == max(Count))]
        ) %>%
        mutate(
            tip_label = paste0(
                representativeSeq, "_S", individual, "_I", immunisation, "_B", boost
            )
        )

    individualData[[ind]] <- tmpData

    individualCDR3 <- AAStringSet(tmpData$CDR3)
    names(individualCDR3) <- tmpData$tip_label

    if (length(individualCDR3) == 0) {
        message(paste("No CDR3 sequences found for individual", ind))
        next
    }

    if (!requireNamespace("msa", quietly = TRUE)) {
        message(paste("Package 'msa' is not installed. Skipping alignment and tree generation for individual", ind))
        next
    }

    # message("---> multiple alignment")
    msa <- msa::msa(individualCDR3, method = "ClustalOmega", type = "protein")
    alignment <- msa::msaConvert(msa, type = "seqinr::alignment")

    # message("---> tree generation")
    dist <- dist.alignment(alignment)
    smalltree <- njs(dist) # to allow for missing values in the distance matrix

    # message("---> tree metadata")
    treeMetaData <- tmpData[c("tip_label", "sample", "individual", "immunisation", "boost")]

    # message("---> store results")
    treeDataList[[ind]] <- smalltree
    treeMetaDataList[[ind]] <- treeMetaData

    # message("---> combine metadata in plot")
    treePlotBase <- ggtree(smalltree, layout = "rectangular") %<+% treeMetaData

    # message("---> generate tree plot")
    annotatedTree <- treePlotBase +
        geom_tiplab(align = TRUE, linesize = .2, size = 2, offset = -0.8) +
        geom_rootpoint(color = "black", size = 3) +
        geom_tippoint(aes(color = boost), size = 2, alpha = .5)

    treePlotList[[ind]] <- annotatedTree
}

```

```{r}
#| label: setOut
out <- NULL
```

```{r eval=dotree}
#| label: run-summary-md
#| include: false

# Esegue il rendering dei documenti figlio per ciascun individuo, utilizzando i dati dell'evoluzione delle sequenze CDR3.

out <- NULL
for (ind in unique(boostEvol$individual)) {
    env <- new.env()
    out <- c(out, knit_child(loopFile, envir = env))
}

```

`r paste(out, collapse = '\n')`

```{r}
#| label: exportCSV
#| include: false
save.image("nanobodies_report.RData")
# Export CSVs for the interactive Python report
write_csv(clustercounts, "clustercounts.csv")
write_csv(cdrcounts,     "cdrcounts.csv")
write_csv(cdrhists,      "cdrhists.csv")
write_csv(clusterbig,    "clusterbig.csv")
write_csv(fastaSeq,      "fastaSeq.csv")
```

# Additional QC

To be completed.

```{r}
#| label: viewlengthdist
#| eval: false
for (samplename in c("317-D1-Bst3_S1", "317-D1-Bst4_S2", "317-L1-Bst2_S3", "317-L1-Bst3_S4", "317-L1-Bst4_S5")) {
    lengthfreqs <- read.csv(file = paste0(samplename, "_L001.hist"), sep = "\t", header = FALSE)
    colnames(lengthfreqs) <- c("sequence_length", "count")
    print(ggplot(data = lengthfreqs) +
        geom_histogram(aes(x = sequence_length, y = count), stat = "identity") +
        ggtitle(paste0("Distribution of merged sequence lengths for sample ", samplename)))
}
```

# Methods

The overall approach is inspired by the work *Large diversity of Functional Nanobodies from a Camelid Immune Library Revealed by an Alternative Analysis of Next-Generation sequencing data* by Pieter Deschaght and colleagues (doi: 10.3389/fimmu.2017.00420, published April 2017).

## Merging forward and reverse reads

Forward and reverse reads have been merged using the software FLASH, with default settings except maximum overlap between the reads, which was set at 300bp. More information about the software is available at their [website](https://ccb.jhu.edu/software/FLASH/)

## Translating cDNA to protein sequence

Translation has been done with an in-house python script developed by T. Bleazard, and using Biopython. In order to identify the correct frame, we have set the starting sequence as *ATGGCACAG* and the end sequence as *ACCGTCTCCTCA*. Anything in between is translated into a protein sequence. If start or end cannot be identified, or if the sequence in between is not a multiple of codon triplets, the sequence will fail translation.

## Clustering the aminoacid sequences

Clustering of the translated sequences is performed using the software CD-HIT. For more information, please visit their [website](https://github.com/weizhongli/cdhit/wiki). In our implementation we have selected an identity cutoff of 0.9, as per results of Deschaght et al.

## Extracting the CDR3 sequences

The CDR3 is annotated on the cluster representatives with nanoCDR-X (Bagordo et al. 2026), a deep-learning model for nanobody CDR annotation. The reference definition of the CDR3 is CDR3-IMGT (Lefranc et al. 2003): the residues between the second conserved cysteine (IMGT position 104) and the J-region tryptophan (IMGT position 118), both excluded, which is also the `cdr3` field of the AIRR Community standard (Vander Heiden et al. 2018). The interactive report compares every nanoCDR-X call with this definition, locating the two anchors with sequence motifs.
